---
title: "Cotton Data Analysis"
author: "Stephanie Wairagu"
date: '2/1/2023'
output: 
  html_document: 
    theme: cerulean
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Background

The aim of this research project is to explore and analyze cotton data provided by Matthew Starr of the Department of Crop and Soil Sciences at North Carolina State University to understand the effect of a plant-based amino acid biostimulant on cotton.


## Getting Started

We'll use the **tidyverse** package for this analysis. If not previously installed, you can do so by running the following command in the console:

```{r eval=FALSE}
install.packages("tidyverse")
```

Next, run the following library command to use the **tidyverse** package:

```{r message=FALSE, eval=TRUE}
library(tidyverse)
```


# Data

We can now read in our dataset using the following command:

```{r load-data, message=FALSE, eval=TRUE}
cotton_data <- read.csv("data/cotton-data.csv")
```


To view the data in your data viewer, run `View(cotton_data)` in the console. This will pop open a separate window with the data.


This dataset has `r nrow(cotton_data)` rows and `r ncol(cotton_data)` columns.

Each row represents a unique cotton plant observation. \
Each column represents a variable. To get a list of the variables in the data frame, we can use the `names()` function:
```{r}
names(cotton_data)
```


## Exploratory Data Analysis

We first create a visualization for the distribution of the **open boll weight** of cotton plants in both the control and treatment groups: 

```{r}
ggplot(cotton_data, aes(x = Open.Boll.Weight..g. ,
                        fill = Treatment)) +
  geom_histogram(bins = 15) +
  labs(x = "Open boll weight (g)",
       y = "Frequency",
       title = "Distribution of the Open Boll Weight",
       subtitle = "By Treatment Group") +
  facet_wrap(~ Treatment) +
  theme(legend.position = "none")
```

After examining the distribution of the open boll weight for the control and treatments groups, it was found that the distribution for the treatment group is *right-skewed* and the sample size is only 20 (not large enough for the Central Limit Theorem to take effect if using a t-test to compare the means of the two groups).





