---
title: "Soybean Data Analysis"
author: "Stephanie Wairagu"
date: '2023-05-05'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(infer)
```

```{r load-data, message=FALSE, eval=TRUE}
soybean <- read_csv("data/soybean-data.csv")
```

To view the data in the data viewer, run `view(soybean)` in the console. This will pop open a separate window with the data. The dataset has `r nrow(soybean)` rows and `r ncol(soybean)` columns. Each row represents a unique soybean plant, treatment observation. Each column represents a variable. 

```{r}
names(soybean)
```

```{r rename-variables}
soybean <- soybean %>% 
  rename(
    Sample_ID = `Sample ID`,
    Treatment_group = `Group (iQ/Control)`,
    Num_of_Plants = `# of Plants`,
    Num_of_Pods = `# of Pods`,
    Num_of_Seeds = `# of Seeds`,
    Seed_Weight_g = `Weight of Seeds (g)`,
    Num_of_Pods_pp = `# of Pods / Plant`,
    Num_of_Seeds_pp = `# of Seeds / Plant`,
    Seed_Weight_pp = `Weight of Seeds / Plant (g)`,
    Weight_per_100_seeds = `Weight per 100 seeds`,
    Oil_content = `Seed Oil Content (%)`,
    Protein_content = `Seed Protein Content (%)`
    )
```

```{r}
names(soybean)
```

### Response: Number of Pods

```{r}
bootstrap_dist_pods <- soybean %>% 
  specify(formula = Num_of_Pods ~ Treatment_group) %>% 
  generate(reps = 1000, type = "bootstrap") %>% 
  calculate(stat = "diff in means", order = c("iQ", "Control"))
```

```{r}
percentile_ci_pods <- bootstrap_dist_pods %>% 
  get_confidence_interval(level = 0.95, type = "percentile")
percentile_ci_pods
```

```{r}
visualize(bootstrap_dist_pods) + 
  shade_confidence_interval(endpoints = percentile_ci_pods)
```

```{r}
obs_diff_means_pods <- soybean %>% 
  group_by(Treatment_group) %>% 
  summarize(mean_pods = mean(Num_of_Pods)) %>% 
  summarize(diff_means = diff(mean_pods)) %>% 
  pull()
```

```{r}
se_ci <- bootstrap_dist_pods %>% 
  get_confidence_interval(level = 0.95, type = "se", 
                          point_estimate = obs_diff_means_pods)
se_ci
```

```{r}
visualize(bootstrap_dist_pods) + 
  shade_confidence_interval(endpoints = se_ci)
```

### Response: Number of Seeds

```{r}
bootstrap_dist_seeds <- soybean %>% 
  specify(formula = Num_of_Seeds ~ Treatment_group) %>% 
  generate(reps = 1000, type = "bootstrap") %>% 
  calculate(stat = "diff in means", order = c("iQ", "Control"))
```

```{r}
percentile_ci_seeds <- bootstrap_dist_seeds %>% 
  get_confidence_interval(level = 0.95, type = "percentile")
percentile_ci_seeds
```

```{r}
visualize(bootstrap_dist_seeds) + 
  shade_confidence_interval(endpoints = percentile_ci_seeds)
```

### Response: Oil Content

```{r}
bootstrap_dist_oil <- soybean %>% 
  specify(formula = Oil_content ~ Treatment_group) %>% 
  generate(reps = 1000, type = "bootstrap") %>% 
  calculate(stat = "diff in means", order = c("iQ", "Control"))
```

```{r}
percentile_ci_oil <- bootstrap_dist_oil %>% 
  get_confidence_interval(level = 0.95, type = "percentile")
percentile_ci_oil
```

```{r}
visualize(bootstrap_dist_oil) + 
  shade_confidence_interval(endpoints = percentile_ci_oil)
```


### Response: Protein Content

```{r}
bootstrap_dist_protein <- soybean %>% 
  specify(formula = Protein_content ~ Treatment_group) %>% 
  generate(reps = 1000, type = "bootstrap") %>% 
  calculate(stat = "diff in means", order = c("iQ", "Control"))
```

```{r}
percentile_ci_protein <- bootstrap_dist_protein %>% 
  get_confidence_interval(level = 0.95, type = "percentile")
percentile_ci_protein
```

```{r}
visualize(bootstrap_dist_protein) + 
  shade_confidence_interval(endpoints = percentile_ci_protein)
```















